Solving the maximum independant set problem by generic programming einsum networks \thanks{\funding{...}}

### 2021-08-06 paper/paper.tex
update

\documentclass[review,onefignum,onetabnum]{siamart190516}
%\usepackage{natbib}
\pdfoutput=1
%\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithmic}
\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

@@

\section{Introduction}
MIS problem is hard~\cite{Hastad1996}.
Branching algorithms can solve the MIS problem in $1.1893^n n ^{O(1)}$ time ~\cite{Xiao2017}.
Previous dynamic programming approach~\cite{Fomin2013} can reduce the complexity of computing to $O(2^{tw(G)}tw(G)n)$.
A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
It is related to hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.

In this work, we attack this problem by mapping it to an ``einsum'' network.
The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of multiplication between two matrices to the contraction between multiple tensors.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
It is denoted as $C_i^k = A_i^j B_j^k$ in the Einstein's original notation, where the paired subscript and superscript $j$ is a dummy index summed over.
An example of tensor networks is $C_i^l = A_{ij}^k B^l_k V^j$.
One can map a tensor network to a muti-graph with open edges by viewing a tensor in the expression on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining labels as open edges.
We get the graphical notation as the following.

@@

One can easily check a label in a tensor network representation appears precisely twice.
Numpy programmers make a generalization to this notation by not restricting the number of times a label is used by tensors.
For example, $C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is an einsum but not a tensor network. Here, all indices not appearing in the output are summed over, i.e. it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Whether the index appear as a superscript or a subscript makes no sense now.
The graphical representation of an einsum is a hypergraph, where an edge can be shared by a arbituary number of nodes.

\centerline{\begin{tikzpicture}
    \def\dx{0};
    \def\r{0.25}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \filldraw[fill=black] (\ax+\dx,\ax) circle [radius=\r];
    \filldraw[fill=black] (\bx+\dx,\by) circle [radius=\r];
    \filldraw[fill=black] (\cx+\dx,\cy) circle [radius=\r];
    \draw[color=blue,thick] (\bx+\dx,\by) circle [radius=\r+0.1];
    \node[color=blue] at (\bx+\dx-0.6,\bx) {i};
    \draw[color=cyan,dashed,thick] (\bx+\dx,\by) circle [radius=\r+0.25];
    \node[color=cyan] at (\bx+\dx+0.5,\bx-0.5) {l};
    \draw[color=violet,thick,dashed] (\ax+\dx,\ay) circle [radius=\r+0.1];
    \node[color=violet] at (\ax+\dx+0.4,-0.4) {k};
    \draw[color=black,thick] (0.5+\dx,0) circle [radius=1.7];
    \node[color=black] at (0.5+\dx-1.9,0) {m};
    \draw[color=red,thick,rotate=135,dashed] (-0.6,0) ellipse (1.2 and 0.5);
    \node[color=red] at (\ax+\dx,-1.0) {j};
    \node[color=white] at (\ax+\dx,\ax) {A};
    \node[color=white] at (\bx+\dx,\by) {B};
    \node[color=white] at (\cx+\dx,\cy) {V};
\end{tikzpicture}}

In the main text, we stick to the einsum notation rather than the tensor network notation,
although one can easily translate an einsum network to the equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
We do not use the language of tensor network because it can sometime increase the contraction complexity of a graph. We will show an example in the appendix.

\section{Independence polynomial}
\begin{figure*}[h!]
    \centerline{\begin{tikzpicture}
        \def\dx{0};
        \def\r{0.25}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25}
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=black,fill=white] at (0.75+\dx,0) {\scriptsize $B$};
        \node[color=black,fill=white] at (\dx,0.75) {\scriptsize $B$};
        \node[color=black,fill=white] at (1.5+\dx,0.75) {\scriptsize $B$};
        \node[color=black,fill=white] at (0.75+\dx,1.5) {\scriptsize $B$};
        \node[color=black,fill=white] at (0.75+\dx,0.75) {\scriptsize $B$};
        \node[color=black,fill=white] at (2+\dx,2) {\scriptsize $B$};
        \node[color=black,fill=white] at (\dx,0) {\scriptsize $W(x_a)$};
        \node[color=black,fill=white] at (\dx,1.5) {\scriptsize $W(x_b)$};
        \node[color=black,fill=white] at (1.5+\dx,0) {\scriptsize $W(x_c)$};
        \node[color=black,fill=white] at (1.5+\dx,1.5) {\scriptsize $W(x_d)$};
        \node[color=black,fill=white] at (2.5+\dx,2.5) {\scriptsize $W(x_e)$};
    \end{tikzpicture}}
    \caption{Mapping a graph to an einsum network.}\label{fig:einsummapping}
\end{figure*}

Let us map the graph $G$ into an einsum network, as shown in \Fig{fig:einsummapping}, by placing a rank one tensor of size $2$ on vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},
\end{equation}
and a rank two tensor of size $2 \times 2$ on edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},
\end{equation}
where a tensor index $s_i$ is a boolean variable that being 1 if vertex $i$ is in the independent set, 0 otherwise, $x_i$ is a variable.
We denote the contraction result of this einsum network as
\begin{equation}
    A(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum runs over all possible vertex configurations and accumulates the product of tensor elements to the output.
Let $x_i = x$ be the same variable, then the product over vertex tensors provides a factor $x^k$, where $k=\sum_i s_i$ is the vertex set size, 
and the product over edge tensors provides a factor $0$ for configurations not being an independent set.
The contraction of this einsum network gives the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
By mapping the independence polynomial solving problem to the einsum network contraction, one can take the advantage of recently developed techiniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$, where $tw(G)$ is the treewidth of $G$.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize fast basic linear algebra subprograms (BLAS) functions for certain tensor element types.

@@

In the rest of this paper, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the einsum network programming generic~\cite{Stepanov2014}.

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).
    \end{split}
\end{align}
By contracting the einsum network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem.
The only problem of this method is it suffers from a space overhead that propotional to the maximum independant set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of $m+1$ random real numbers, e.g. $\{0, 1, 2, \ldots, m\}$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is esentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).
\end{align}

@@

\subsection{The finite field algebra approach}
It is possible to compute the independence polynomials exactly using 64 bit integers types only,
even when the factors are larger than that can be represented by 64 bit integers.
We achieve this by designing a finite field algebra $GF(p)$

@@

In a finite field algebra, we have the following observations
\begin{enumerate}
    \item One can still use Gaussian elimination~\cite{Golub2013} to solve a linear equation.
    This is because a field has the property that the multiplicative inverse exists for any non-zero value.
    The multiplicative inverse here can be computed with the extended Euclidean algorithm.
    \item Given the remainders of a larger integer $x$ over a set of coprime integers $\{p_1, p_2, \ldots, p_n\}$,
    $x \pmod {p_1 \times p_2 \times \ldots \times p_n}$ can be computed using the chinese remainder theorem.
    With this, one can infer big integers even though its bit width is larger than the register size.
\end{enumerate}
With these observations, we developed Algorithm~\ref{alg:finitefield} to compute independent polynomial exactly without introducing space overheads.
In the algorithm, except the computation of chinese remainder theorem, all computations are done with integers with fixed width $W$.

@@

We do the same computation as independence polynoimal, the coefficients of resulting polynomial gives the counting of maximal independent sets.
In many sparse graphs, this tensor network contraction approach is much faster than computing the maximal cliques of its complement and use Bron Kerbosch algorithms for finding maximum cliques.
However, the treewidth of this new tensor network is larger than the one for independence polynomial because it can not utilize some structures of the original graph,
while the original tensor network can be trivially reduced to this one.
We will use an example in the appendix to show why this tensor network is harder to contract.

@@

We can

\section{benchmarks}
We run a sequetial program benchmark on CPU Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz, and show the results bellow.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{benchmark.pdf}
    \caption{Benchmark results for computing different properties with different element types.
    }\label{fig:benchmark}
\end{figure}

>>>>

\documentclass[review,onefignum,onetabnum]{siamart190516}
%\usepackage{natbib}
%\usepackage[sort]{cite}
\pdfoutput=1
%\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usetikzlibrary{snakes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithmic}
\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

\newcounter{example}
\newenvironment{example}[1][]{\refstepcounter{example}\par\medskip
   \noindent \textbf{Example~\theexample. #1} \rmfamily}{\medskip}

@@

\section{Introduction}
In this work, we introduce a new method to solve the classic graph problem, finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
Finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbituary small possitive $\epsilon$.
The naive algorithm of enumerating all configuration space gives a $2^{|V|}$ time solution.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm gives a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving this problem better not only because it is a NP-complete problem that directly related to other NP-complete prolems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
In these applications, knowing the MIS size and one of the optimal solution is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all indenepent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements.

\section{Einsum network}
The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of multiplication between two matrices to the contraction between multiple tensors.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
It is denoted as $C_i^k = A_i^j B_j^k$ in the Einstein's original notation, where the paired subscript and superscript $j$ is a dummy index summed over.
One can map a tensor network to a muti-graph with open edges by viewing a tensor in the expression on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^k B^l_k V^j$, where its graphical notation as the following.
One can easily check a label in a tensor network representation appears precisely twice.

@@

\end{example}

Numpy programmers make a generalization of this notation by not restricting the number of times a label is used by tensors,
hence whether an index appears as a superscript or a subscript makes no sense now.
It has different names in different context, like sum-product network and factor graph~\cite{Bishop2006}.
The graphical representation of an einsum is a hypergraph, where an edge (degree of freedom) can be shared by a arbituary number of nodes (tensors).

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is an einsum but not a tensor network, it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Its hypergraph representation is as the following, where we use different color to annotate different hyperedges.

\centerline{\begin{tikzpicture}[
    dot/.style = {circle, fill, minimum size=#1,
                inner sep=0pt, outer sep=0pt},
    dot/.default = 6pt  % size of the circle diameter 
                    ]  
    \def\dx{0};
    \def\r{0.5cm}
    \def\ax{0}
    \def\ay{0}
    \def\bx{1}
    \def\by{1}
    \def\cx{1}
    \def\cy{-1}
    \node[color=white,fill=black,dot=\r] at (\ax+\dx,\ax) (a) {A};
    \node[color=white,fill=black,dot=\r] at (\bx+\dx,\by) (b) {B};
    \node[color=white,fill=black,dot=\r] at (\cx+\dx,\cy) (v) {V};
    \node at (\ax-0.8,\ay) (k) {k};
    \node at (\bx+0.4,\ay) (m) {m};
    \node at (\ax,\cy) (j) {j};
    \node at (\bx+1,\by) (l) {l};
    \node at (\bx-1,\by) (i) {i};
    \draw[color=blue] (i) -- (b);
    \draw[color=blue] (i) -- (\ax-1,\by);
    \draw[color=cyan,thick] (l) -- (b);
    \draw[color=violet,thick] (k) -- (a);
    \draw[color=violet,thick] (k) -- (\ax-1.5,\ay);
    \draw[color=black,thick] (b) -- (m);
    \draw[color=black,thick] (m) -- (a);
    \draw[color=black,thick] (m) -- (v);
    \draw[color=red,thick] (a) -- (j);
    \draw[color=red,thick] (v) -- (j);
    \draw[color=red,thick] (\ax-1,\cy) -- (j);
\end{tikzpicture}}
\end{example}

In the main text, we stick to the einsum notation rather than the tensor network notation.
As a note to those who are more familiar with tensor network representation,
although one can easily translate an einsum network to the equivalent tensor network by adding $\delta$ tensors (a generalization of identity matrix to higher order).
It can sometime increase the contraction complexity of a graph. We have an example demonstrating this in \App{app:tensorbad}.

\section{Independence polynomial}
One can encode the independent set problem on graph $G$ to an einsum network by placing a rank one tensor of size $2$ on vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
$x_i$ is a variable.
The contraction of such an einsum network gives
\begin{equation}
    A(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output.
Let $x_i = x$, then the product over vertex tensors gives a factor $x^k$, where $k=\sum_i s_i$ is the vertex set size, 
and the product over edge tensors gives a factor $0$ for configurations not being an independent set.
The contraction of this einsum network gives the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
By mapping the independence polynomial solving problem to the einsum network contraction, one can take the advantage of recently developed techiniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$, where $tw(G)$ is the treewidth of the line graph of a tensor network, here it corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize fast basic linear algebra subprograms (BLAS) functions for certain tensor element types.

\begin{example}
    Mapping a graph (left) to an einsum network, the resulting einsum network is shown in the right panel.
    A vertex is mapped to a hyperedge in the einsum's graphical notation.
    An edge is mapped to an edge tensor.

    \centerline{\begin{tikzpicture}[
    dot/.style = {circle, fill, minimum size=#1,
                inner sep=0pt, outer sep=0pt},
    dot/.default = 6pt  % size of the circle diameter 
                    ]

        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}
    \begin{align*}
        \sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}.
    \end{align*}
\end{example}

@@

In the rest of this paper, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the einsum network generic~\cite{Stepanov2014}.

\subsection{The polynomial approach}
A straight forward approach to evaluate the independence polynomial is treating the tensor elements as polynomials, and evaluate the polynomial directly.
Let us create a polynomial type, and represent a polynomial $a_0 + a_1 x + \ldots + a_k x^k$ as a vector $(a_0, a_1, \ldots, a_k) \in R^k$, e.g. $x$ is represented as $(0, 1)$.
We define the algebra between the polynomials $a$ of order $k_a$ and $b$ of order $k_b$ as
\begin{align}
    \begin{split}
    a \oplus b &= (a_0 + b_0, a_1 + b_1, \ldots, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_0 + b_0, a_1b_0 + a_0b_1, \ldots, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (),\\
    \mymathbb{1} &= (1).\label{eq:polynomial}
    \end{split}
\end{align}
By contracting the einsum network with polynomial type, the final result is the exact representation of the independence polynomial.
In the program, the multiplication can be evaluated efficiently with the convolution theorem.
The only problem of this method is it suffers from a space overhead that propotional to the maximum independant set size because each polynomial requires a vector of such size to store the factors.
In the following subsections, we managed to solve this problem.

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of $m+1$ random real numbers, e.g. $\{0, 1, 2, \ldots, m\}$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is esentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

@@

\subsection{The finite field algebra approach}
It sounds a bit over ambitious to compute the independence polynomial regorously using integer number types only,
because the fixed width integer types are often too small to store the countings,
while big integer with arbituary precision can be very slow and imcompatible with graphic processing units (GPU) devices.
The solution we found is to computation on a finite field algebra $GF(p)$

@@

In a finite field algebra, we have the following observations
\begin{enumerate}
    \item One can still use Gaussian elimination~\cite{Golub2013} to solve a linear equation \Eq{eq:lineareq}.
    This is because a field has the property that the multiplicative inverse exists for any non-zero value.
    The multiplicative inverse here can be computed with the extended Euclidean algorithm.
    \item Given the remainders of a larger integer $x$ over a set of coprime integers $\{p_1, p_2, \ldots, p_n\}$,
    $x \pmod {p_1 \times p_2 \times \ldots \times p_n}$ can be computed using the chinese remainder theorem.
    With this, one can infer big integers even though its bit width is larger than the register size.
\end{enumerate}
With these observations, we developed Algorithm~\ref{alg:finitefield} to compute independence polynomial exactly without introducing space overheads.
In the algorithm, except the computation of chinese remainder theorem, all computations are done with integers of fixed width $W$.

@@

We do the same computation as independence polynomial, the coefficients of resulting polynomial gives the counting of maximal independent sets.
In many sparse graphs, this tensor network contraction approach is much faster than computing the maximal cliques of its complement and use Bron Kerbosch algorithms for finding maximum cliques.
However, the treewidth of this new tensor network is larger than the one for independence polynomial because it can not utilize some structures of the original graph,
while the original tensor network can be trivially reduced to this one.
We will use an example in the appendix to show why this tensor network is harder to contract.

@@

We can see these two subgraphs produce exactly the same tensors.

\section{benchmarks}
We run a sequetial program benchmark on CPU Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz, and show the results bellow.
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{benchmark.pdf}
    \caption{Benchmark results for computing different properties with different element types.
    }\label{fig:benchmark}
\end{figure}
Einsum network contraction is parallelizable. When the element type is immutable, one can just upload the data to GPU to enjoy the speed up.
